On the analytic structure of the linear response susceptibility in open systems

### 2018-04-16 AnalyticStructure_main.tex
update

As an explicit example in support of these arguments we compare the absorption spectrum of a $CO$ molecule computed in three distinct real-space basis set, which differs for the dimension of 
the computational domain. Results reported in figure \ref{co_spectrum} evidences that the curves are almost coincident in the below threshold regime, whereas the effect of the computational 
setup are clearly visible in the for values of $\omega$ that exceed the threshold value.

@@

%This analysis, carried out starting from the representation of the linear response functional in terms of fluctuation states, evidences some general properties of the linear response. 
%In the next sections we will reformulate an analogous procedure directly looking at the susceptibility. This analysis lead us to the core of the paper, namely the description of the
%analytical structure of the linear response susceptibility.

@@

The localization properties of the response density can be reformulated in terms of the \emph{susceptibility functional}, implicitly defined trough the operator equation  
\be\lb{SusceptibilityDef1}
\dm'(\omega) = \sop{{\cal R}}(\omega) \op\Phi(\omega) \;. 
\ee
The main advantage of this approach relies on the fact that the susceptibility describes an intrinsic feature of the unperturbed system and its knowledge determines the LR properties 
independently from the specific choice of the perturbing field.

The analysis is based on the spectral decomposition of the Liouville equation \eqref{LiouvillianRhopomegaDef1} and the make usage of the \emph{excitation operators}, defined through the 
left and right action eigenvalue equations 
\be\lb{ExcitationOperatorsDef1}
\Liouv \op E = E \op E \qq \op{\tilde E} \Liouv = E \op{\tilde E} \;,
\ee
with the normalization condition $\trace{\op{\tilde E}\op E'} = \delta(E-E')$. The properties of the Liouvillian superoperator (see appendix?) imply that excitations are transverse 
operators so that $\op E = \dmnot\op E\op Q_0 + \op Q_0\op E\dmnot$. This condition allows us to introduce a dyadic representation of excitations as follows
\begin{align}\lb{ExcitationOperatorsDef2}
\op E &= \sum_p \left( \ket{\phi^E_p}\bra{\psi_p} + \ket{\psi_p} \bra{\tilde\chi^E_p}\right)\;, \nn \\
\op{\tilde E} &= \sum_p \left( \ket{\chi^E_p}\bra{\psi_p} + \ket{\psi_p} \bra{\tilde\phi^E_p}\right)\;,
\end{align}
and the \emph{excited states} $\ket{\phi^E_p}$ and $\bra{\tilde\chi^E_p}$ and their conjugates span the domain of $\op Q_0$, \emph{i.e} the empty subspace of $\hnot$.

We restrict our analysis assuming that a set of hypothesis commonly realized in DFT holds (symmetry of the transition basis, specify better? see appendix?). In this case excitation
eigenvalues come in pair. It is possible to show that and if a couple of left and right eigenstates satisfy \eqref{ExcitationOperatorsDef1} with energy $E$ than their transpose 
%$(\op E,\op{\tilde E})$ $(\op E^t,\op{\tilde E}^t)$ 
are eigenstates of the same equations with eigenvalue $-E$. Equations \eqref{ExcitationOperatorsDef1} can be recasted in the form of a Sternheimer-like equations for the excited states. 
In particular, the left action eigenvalue equation provides
%We further analyze the first equation of \eqref{ExcitationOperatorsDef1} under a.   
%In this case the excitation eigenvalues come in pair and for each 
%that can be recasted in the form of a Sternheimer-like equations for the excited states. We assume that
%the condition $(\op V'[\op E])^t=\op V'[\op E^t]$
%The eigenvalue 
%condition \eqref{ExcitationOperatorsDef1} 
%can thus be recasted in the 
% \begin{align}
% &\left(\hnot - \eps_p\right) \ket{\phi_p^E} + \op Q_0 \op V'[\op E] \ket{\psi_p} = E \ket{\phi_p^E} \;, \nn\\
% &\left(\hnot - \eps_p\right)\ket{\tilde\chi_p^E} + \op Q_0\op V'[\op E^t]\ket{\psi_p}   = -E\ket{\tilde\chi_p^E}  \;, 
% \end{align}
\begin{align}\lb{ExcitationOperatorsDef3}
&\left[E-(\hnot - \eps_p)\right] \ket{\phi_p^E} = \op Q_0 \op V'[\op E] \ket{\psi_p} \;, \nn\\
&\left[-E-(\hnot - \eps_p)\right]\ket{\tilde\chi_p^E} = \op Q_0\op V'[\op E^t]\ket{\psi_p}  \;, 
\end{align}
and the orthonormalization condition of the excitation operators is expressed as 
\be\lb{ExcitedStateOrthNormDef1}
\sum_p \left(\brket{\tilde\phi_p^E}{\phi_p^{E'}} + \brket{\tilde\chi_p^{E'}}{\chi_p^E}\right) = \delta(E-E') \;. 
\ee
Equations \eqref{ExcitationOperatorsDef3} allows us to present a localization argument for the excited states using the same procedure adopted for fluctuation states in 
section \ref{FluctuationState}. Pursing this strategy we present a formal solution of \eqref{ExcitationOperatorsDef3} as follows 
\begin{align}
&\ket{\phi^E_p} = \op G_H(E+\eps_p)\ket{s_{\phi}} \;, \nn \\
&\ket{\tilde\chi^E_p} = \op G_H(-E+\eps_p)\ket{s_{\tilde\chi}} \;,
\end{align}
where $\op G_H$ is the Helmholtz Green function and the (self-consistent) source term $s_\phi$ and $s_{\tilde\chi}$ collect the sum of the r.h.s of \eqref{ExcitationOperatorsDef3} 
plus the action of the KS potential on the excited states.

We focus our analysis on solutions with positive energy, \emph{i.e.} with $E>0$. Looking at the coordinate representation of $\op G_H$ (see eq. \eqref{HelmholtzKernelDef1}) we conclude
that the Helmholtz kernel evaluated in $-E+\eps_p$ contains an exponential damping factor for positive values of $E$. As a consequence all the $\ket{\tilde\chi^E_p}$ states
exhibit a bound-state behavior in the long range. Instead, the asymptotic character of $\op G_H(E+\eps_p)$ depends from the value of $E$ and $\ket{\phi^E_p}$ varies from a bound-state 
like behavior for $E<|\eps_p|$, to a delocalized spherical-wave like character for values of $E>|\eps_p|$. 
%On the contrary, the localization character of $\ket{\phi^E_p}$ varies from bound-state like solutions, for $E<|\eps_p|$, to delocalized spherical-wave like 
%behavior for values of $E>|\eps_p|$. 
There is however a novel aspect which did not emerged in the analysis of fluctuation states and is determined by the presence of the further condition \eqref{ExcitedStateOrthNormDef1}.
%The consequence of this condition 
To understand this point we observe that Sternheimer equations \eqref{ExcitationOperatorsDef3} are the first order expansion of a set of Hamiltonian eigenvalue equations. It is well known 
that the normalization constraints have the effect of determining the quantization of accessible energy levels for bound-state solutions. 
Excited states with $E<|\eps_p|$ constitute a genuine \emph{discrete} set while the ones in the unbound energy range $E>|\eps_p|$ belong to a continuum.

This considerations determine the structure of the excitations spectrum and essentially evidence that there exists two distinct sectors of the exctiation spectrum. The first one is composed by
localized excitations built in terms of excited states below their threshold level. This set has a finite number of elements.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[t]
\begin{tikzpicture}[scale=0.8]
% axes
\draw[->] (0,0) -- (8,0); 
\draw[->] (0,0) -- (0,6);
% orizontal lines
\draw[dashed,thick,color=gray] (0,5) -- (2,5);
\draw[very thick](2,5) -- (8,5);
\draw[dashed,thick,color=gray] (0,3) -- (4,3);
\draw[very thick] (4,3) -- (8,3);
\draw[dashed,thick,color=gray] (0,1) -- (6,1);
\draw[very thick] (6,1) -- (8,1);
% % vertical lines
\draw[dashed,very thin,color=black] (2,0) -- (2,5);
\draw[dashed,very thin,color=black] (4,0) -- (4,3);
\draw[dashed,very thin,color=black] (6,0) -- (6,1);
% % threshold levels labels
\node[left] at (0,5) {$h$};
\node[left] at (0,3) {$i$};
\node[left] at (0,1) {$1$};
\node[below] at (2,0){$|\eps_h|$};
\node[below] at (4,0){$|\eps_i|$};
\node[below] at (6,0){$|\eps_1|$};
\node[below] at (4,-0.5) {Excitation energy};
% bars at thresholds
%\foreach \Point in {(2,5),(4,3),(6,1)}{
%   \node at \Point {{\textbar}};%$\times$};
%}
\draw[very thick] (2,4.9) -- (2,5.1);
\draw[very thick] (4,2.9) -- (4,3.1);
\draw[very thick] (6,0.9) -- (6,1.1);
% dotted vertical lines
%\draw[dotted,thick,color=black] (-0.2,3.6) -- (-0.2,4.4);
%\draw[dotted,thick,color=black] (-0.2,1.6) -- (-0.2,2.4);
% discrete excitations
\foreach \Point in {(0.5,5),(1.0,5),(1.8,5),(0.8,3),(1.3,3),(1.8,3),(2.4,3),(2.8,3),(3.4,3),(3.8,3),(0.7,1),(1.5,1),(1.9,1),(2.6,1),(2.9,1),(3.5,1),(4.1,1),(4.7,1),(5.7,1)}{
    \node at \Point {\textbullet};
}
\end{tikzpicture}
\caption{\label{ExcitationLandscape} Excitation landscape.}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

As an explicit example in support of these arguments we compare the absorption spectrum of a $CO$ molecule computed in three distinct real-space basis set, which differs for the dimension of 
the computational domain. Results reported in figure \ref{co_spectrum} evidences that the curves are almost coincident in the below threshold regime, whereas the effect of the computational 
setup are clearly visible for values of $\omega$ that exceed the threshold value.

@@

@@

The study of the localization properties of the LR functional, formerly discussed through the inspection of the response density operator, can be reformulated in terms of the 
\emph{susceptibility functional}, defined trough the operator equation  
%The localization properties of the response density can be reformulated in terms of the \emph{susceptibility functional}, implicitly defined trough the operator equation  
\be\lb{SusceptibilityDef1}
\dm'(\omega) = \sop{{\cal R}}(\omega) \op\Phi(\omega) \;. 
\ee
This reformulation of the problem has the advantage of providing some general information on the nature of the excitation spectrum of the system. Results achieved in this way thus 
describes an intrinsic feature of the system and determine the properties of the LR independently from the specific choice of the perturbing field. 
%The main advantage of this approach relies on the fact that the susceptibility describes an intrinsic feature of the unperturbed system and its knowledge determines the LR properties 
%independently from the specific choice of the perturbing field.

The analysis is based on the spectral decomposition of the Liouville equation \eqref{LiouvillianRhopomegaDef1} and the make usage of the \emph{excitation operators}, defined through the 
left and right action eigenvalue equations 
\be\lb{ExcitationOperatorsDef1}
\Liouv \op E = E \op E \qq \op{\tilde E} \Liouv = E \op{\tilde E} \;,
\ee
together with the operator orthonormalization condition %$\trace{\op{\tilde E}\op E'} = \delta(E-E')$. 
\be\lb{orthoExcitatioOpDef1}
\trace{\op{\tilde E}\op E'} = \delta(E-E')
\ee
The properties of the Liouvillian superoperator (see appendix?) imply that excitations are transverse 
operators so that $\op E = \dmnot\op E\op Q_0 + \op Q_0\op E\dmnot$. This condition allows us to introduce a dyadic representation of excitations as follows
\begin{align}\lb{ExcitationOperatorsDef2}
\op E &= \sum_p \left( \ket{\phi^E_p}\bra{\psi_p} + \ket{\psi_p} \bra{\tilde\chi^E_p}\right)\;, \nn \\
\op{\tilde E} &= \sum_p \left( \ket{\chi^E_p}\bra{\psi_p} + \ket{\psi_p} \bra{\tilde\phi^E_p}\right)\;,
\end{align}
where both the \emph{excited states} $\ket{\phi^E_p}$, $\bra{\tilde\chi^E_p}$ and their conjugates span the empty subspace of $\hnot$.

We restrict our analysis assuming that a set of hypothesis commonly realized in DFT holds (symmetry of the transition basis, specify better? see appendix?). 
%In this case excitation eigenvalues come in pair. It is possible to show that and if a couple of left and right eigenstates satisfy \eqref{ExcitationOperatorsDef1} with energy $E$ 
%than their transpose %$(\op E,\op{\tilde E})$ $(\op E^t,\op{\tilde E}^t)$ 
%are eigenstates of the same equations with eigenvalue $-E$. 
Equations \eqref{ExcitationOperatorsDef1} can be recasted in the form of a Sternheimer-like equations for the excited states. 
In particular, the left action eigenvalue equation provides   
\begin{align}\lb{ExcitationOperatorsDef3}
&\left[E-(\hnot - \eps_p)\right] \ket{\phi_p^E} = \op Q_0 \op V'[\op E] \ket{\psi_p} \;, \nn\\
&\left[-E-(\hnot - \eps_p)\right]\ket{\tilde\chi_p^E} = \op Q_0\op V'[\op E^t]\ket{\psi_p}  \;, 
\end{align}
and the orthonormalization condition \eqref{orthoExcitatioOpDef1} implies %among the excitation operators is expressed as 
\be\lb{ExcitedStateOrthNormDef1}
\sum_p \left(\brket{\tilde\phi_p^E}{\phi_p^{E'}} + \brket{\tilde\chi_p^{E'}}{\chi_p^E}\right) = \delta(E-E') \;. 
\ee
Equations \eqref{ExcitationOperatorsDef3} allows us to present a localization argument for the excited states using the same procedure adopted for fluctuation states in 
section \ref{FluctuationState}. Pursing this strategy we present a formal solution of \eqref{ExcitationOperatorsDef3} as follows 
\begin{align}
&\ket{\phi^E_p} = \op G_H(E+\eps_p)\ket{s_{\phi}} \;, \nn \\
&\ket{\tilde\chi^E_p} = \op G_H(-E+\eps_p)\ket{s_{\tilde\chi}} \;,
\end{align}
where $\op G_H$ is the Helmholtz Green function, while the (self-consistent) source term $s_\phi$ and $s_{\tilde\chi}$ collect the sum of the r.h.s of \eqref{ExcitationOperatorsDef3} 
plus the action of the KS potential on the excited states.

It is possible to show that excitation eigenvalues come in pair and we focus our analysis on solutions with positive energy, \emph{i.e.} with $E>0$. Looking at the coordinate representation 
of $\op G_H$ (see eq. \eqref{HelmholtzKernelDef1}) we observe that the Helmholtz kernel evaluated in $-E+\eps_p$ contains an exponential damping factor for \emph{all} the positive values of $E$. 
As a consequence all the $\ket{\tilde\chi^E_p}$ states exhibit a bound-state behavior in the long range. Instead, the asymptotic character of $\op G_H(E+\eps_p)$ depends from $E$ and we can 
identify the threshold value $E=|\eps_p|$ in terms of which $\ket{\phi^E_p}$ changes from a bound-state like behavior to a delocalized spherical-wave like character.  
Furthermore, the orthonormalization conditions \eqref{ExcitedStateOrthNormDef1} act as a ulterior constraint for the bound-state solutions and have the effect of determining the quantization of 
the corresponding energy levels\footnote{Sternheimer equations represent the first order expansion of a set of Hamiltonian eigenvalue equations. Consequently the properties of their solution...}.
We conclude that all the states $\ket{\tilde\chi^E_p}$ and the $\ket{\phi^E_p}$ with $E$ below the threshold value constitute a genuine \emph{discrete} set, while the ones in the unbound 
energy range behaves as generalized eigenvectors associated to a continuum spectrum.

%To complete the analysis we have to take into account the orthonormalization constraints \eqref{ExcitedStateOrthNormDef1}. Sternheimer equations \eqref{ExcitationOperatorsDef3} are the 
%first order expansion of a set of Hamiltonian eigenvalue equations.  
%Moreover, there is a further consequence due to the orthonormalization condition \eqref{ExcitedStateOrthNormDef1}.
%There is however a novel aspect which did not emerged in the analysis of fluctuation states and is 
%To understand this point we observe that Sternheimer equations \eqref{ExcitationOperatorsDef3} are the first order expansion of a set of Hamiltonian eigenvalue equations. It is well known 
%that the normalization constraints have the effect of determining the quantization of accessible energy levels for bound-state solutions.

This considerations allows us to determine the structure of the excitations spectrum. To achieve this result we assume to have solved the equations \eqref{ExcitedStateOrthNormDef1} for a 
value of $E$.

%Given an excitation operator in the form \eqref{ExcitationOperatorsDef2}

%and essentially evidence that there exists two distinct sectors of the exctiation spectrum. The first one is composed by
%localized excitations built in terms of excited states below their threshold level. This set has a finite number of elements.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[t]
\begin{tikzpicture}[scale=0.78]
% axes
\draw[->] (0,0) -- (8,0); 
\draw[->] (0,0) -- (0,6);
% orizontal lines
\draw[dashed,thick,color=gray] (0,5) -- (2,5);
\draw[very thick](2,5) -- (8,5);
\draw[dashed,thick,color=gray] (0,3) -- (4,3);
\draw[very thick] (4,3) -- (8,3);
\draw[dashed,thick,color=gray] (0,1) -- (6,1);
\draw[very thick] (6,1) -- (8,1);
% % vertical lines
\draw[dashed,very thin,color=black] (2,0) -- (2,5);
\draw[dashed,very thin,color=black] (4,0) -- (4,3);
\draw[dashed,very thin,color=black] (6,0) -- (6,1);
% % threshold levels labels
\node[left] at (0,5) {$l_h$};
\node[left] at (0,3) {$l_i$};
\node[left] at (0,1) {$l_1$};
\node[below] at (2,0){$|\eps_h|$};
\node[below] at (4,0){$|\eps_i|$};
\node[below] at (6,0){$|\eps_1|$};
\node[below] at (4,-0.5) {Excitation energy};
% bars at thresholds
%\foreach \Point in {(2,5),(4,3),(6,1)}{
%   \node at \Point {{\textbar}};%$\times$};
%}
\draw[very thick] (2,4.9) -- (2,5.1);
\draw[very thick] (4,2.9) -- (4,3.1);
\draw[very thick] (6,0.9) -- (6,1.1);
% dotted vertical lines
%\draw[dotted,thick,color=black] (-0.2,3.6) -- (-0.2,4.4);
%\draw[dotted,thick,color=black] (-0.2,1.6) -- (-0.2,2.4);
% discrete excitations
\foreach \Point in {(0.5,5),(1.0,5),(1.8,5),(0.8,3),(1.3,3),(1.8,3),(2.4,3),(2.8,3),(3.4,3),(3.8,3),(0.7,1),(1.5,1),(1.9,1),(2.6,1),(2.9,1),(3.5,1),(4.1,1),(4.7,1),(5.7,1)}{
    \node at \Point {\textbullet};
}
\end{tikzpicture}
\caption{\label{ExcitationLandscape} Excitation landscape.}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
